---
title: "Data exploration assignment"
author: "Devin"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = TRUE)
```

In this exercise, you will practice basic `dplyr` (the relevant package
within `tidyverse`) operations on a dataset of gene expression levels
across human tissues (from the GTEx Project). The dataset (`dt`) has one
row per gene, with columns:

- `Name`: the Ensembl gene ID
- `Description`: the gene symbol
- One column per tissue containing the median expression levels (in TPM
  units)

**Using HopGPT.** You are welcome to use HopGPT on this assignment. If
you do, add a couple of comment lines in that same code block noting
what you asked it and how you checked the answer. Rmember that a chatbot
will happily give you code that runs perfectly and answers a slightly
different question than the one you asked.

For example:

```{r ai-comment-example, eval=FALSE}
# Asked HopGPT how to sort a tibble by a column with a space in its name.
# Checked: printed the first and last rows to confirm it sorted descending.
```

------------------------------------------------------------------------

### Start by loading the data into R:

Note the arguments, which specify that the first two lines of the file
will be skipped, that the file contains a header line that should be
used to form column names, and that the columns are tab (`\t`)
separated.

```{r load-data}
library(tidyverse)

dt <- read_delim("~/Data/GTEx/GTEx_Analysis_2017-06-05_v8_RNASeQCv1.1.9_gene_median_tpm.gct.gz",
                 skip = 2, col_names = TRUE, delim = "\t")
```

### Q1. How many genes are in the dataset, and how many tissues are represented?

Note that not every column is a tissue.

```{r q1}
dt %>%
  select(Name) %>%
  nrow()
dt %>%
  select(-Name, -Description) %>%
  ncol()
# there are 56200 genes in the dataset and 54 tissues are represented

```

------------------------------------------------------------------------

### Q2. How many genes are expressed above a given threshold in each tissue?

A common threshold is 1 TPM to define whether a gene is "expressed."\
Compute, for each of the three tissues, `Liver`, `Testis`, and
`Whole Blood`, how many genes are expressed above this threshold.\
Then identify which tissue has the **most** and which has the **fewest**
expressed genes. See this paper for one possible explanation: Xia et al.
2021: [PMC7891839](https://pmc.ncbi.nlm.nih.gov/articles/PMC7891839/).

Do the results change if you set the threshold higher, say at 10 TPM?

*Hint:* the column name `Whole Blood` contains a space, so it needs
backticks: `` dt$`Whole Blood` ``.

```{r q2}
dt %>%
  filter(Liver > 1) %>%
  nrow()

dt %>%
  filter(Testis > 1) %>%
  nrow()

dt %>%
  filter(`Whole Blood` > 1) %>%
  nrow()

dt %>%
  filter(Liver > 10) %>%
  nrow()

dt %>%
  filter(Testis > 10) %>%
  nrow()

dt %>%
  filter(`Whole Blood` > 10) %>%
  nrow()
# There are 13499 genes expressed above 1 TPM in liver, 25064 in testis, and 11616 in whole blood. Whole blood has the fewest expressed genes and testis has the most. For 10 TPM, there are 5132 genes expressed above that threshold in liver, 11471 in testis, and 4482 in whole blood. So yes, the number of genes expresse din a certain tissue decreases with a higher expression threshold.

```

------------------------------------------------------------------------

### Q2b. Write a function

If you answered Q2 by copying the same line of code and editing the
tissue name, and then copying those lines again and editing the
threshold, you did six near-identical things by hand. Each time you
copied code represents a chance to introduce a mistake, and if you
wanted to change something, you would have to implement the change
multiple times.

Write a function `count_expressed()` that takes a vector of expression
values and a threshold, and **returns** the number of genes above that
threshold. Give the threshold a default value of 1.

Test it on a small vector where you already know the answer, such as
`c(0, 0.5, 1, 5, 100)`. Then use it to redo Q2 (all three tissues, both
thresholds) instead of writing out the counting code separately six
times.

```{r q2b}

# your function here

# test on an input with a known answer

# use it to answer Q2 again

```

------------------------------------------------------------------------

### Q3. What are the 20 most highly expressed genes in the liver?

What do you notice about many of the genes in this list? Can you propose
a hypothesis to explain your observation?

```{r q3}

dt %>%
  arrange(-Liver) %>%
  head(20)

# Most of these genes appear to be mitochondrial genes. To explain this, it's possible that liver cells have a high proportion of mitochondria.

```

------------------------------------------------------------------------

### Q4. Compare the mean expression levels of mitochondrial versus nuclear-encoded genes in the Liver.

Genes encoded by the mitochondrial genome have the gene symbol prefix
`MT-`. The function `str_starts()` from the `stringr` package in
`tidyverse` tests whether a string (or a vector of strings) starts with
a given pattern, returning TRUE/FALSE (or a boolean vector) as output.

Use `str_starts()` to identify which genes are mitochondrially encoded
and store this information in a new Boolean column named
`is_mitochondrial`. Next, compute the mean and median expression of
mitochondrial genes in the liver, as well as the mean and median
expression of nuclear-encoded genes in the liver and compare them. Do
mitochondrial genes exhibit higher or lower mean and median expression
than nuclear genes? What does the difference between the mean and median
tell you about the distribution of gene expression?

**Check your pattern before you trust the result.** There is an easy
version of this that is wrong: matching `"MT"` instead of `"MT-"`. Count
how many genes each of the two patterns matches, and look at a few of
the genes that `"MT"` picks up but `"MT-"` does not. How much does the
mistake change your answer?

```{r q4}

dt %>%
  mutate(is_mitochondrial = str_starts(Description, "MT-")) %>%
  group_by(is_mitochondrial) %>%
  summarize(
    Liver_median = median(Liver, na.rm = TRUE),
    Liver_mean = mean(Liver, na.rm = TRUE))

# how many genes match "MT-" vs "MT"? which genes differ?
dt$Description %>%
  str_starts("MT") %>%
  sum(na.rm = TRUE)

dt$Description %>%
  str_starts("MT-") %>%
  sum(na.rm = TRUE)
# 549 genes match "MT" versus only 37 genes match "MT-"

# It appears that in liver tissue, mitcohondrial genes are expressed at a higher level than nuclear-encoded genes. The high mean for mitochondrial genes but relatively low median suggests that there are a few mitochondrial genes with very high expression that are skewing the mean to a higher value. 
```

------------------------------------------------------------------------

### Q5. Which tissue is this?

So far we have told you which tissue to look at. Now you will find a
tissue based on the genes it expresses.

A gene is *tissue-specific* if it is expressed much more in one tissue
than it is on average across the body. To measure that, you first need
each gene's average expression across all tissues.

```{r mean-across-tissues}
# remove the columns that are not tissue expression
dt <- dt_with_mito
tissue_only <- dt %>%
  select(-Name, -Description, -is_mitochondrial)

# store the mean back in the original tibble
dt <- dt %>%
  mutate(mean_all_tissues = rowMeans(tissue_only))
```

**(a)** Pick any tissue in the dataset and compute a specificity score
for every gene:

$$\text{specificity} = \log_2 \left( \frac{\text{my tissue} + 1}{\text{mean across all tissues} + 1} \right)$$

Keep only genes expressed above 10 TPM in your tissue, sort by
specificity, and print the top 20 gene symbols.

```{r q5a}
#I picked Brain - Amygdala

dt %>%
  filter(`Brain - Amygdala` > 10) %>%
  mutate(specificity = log2((`Brain - Amygdala`+1)/(mean(mean_all_tissues)+1))) %>%
  select(Description,specificity, `Brain - Amygdala`) %>%
  arrange(-specificity) %>%
  select(Description) %>%
  head(20)
  
```

**(b)** Give a classmate *only that list of 20 gene symbols* — not the
tissue name — and see whether they can identify the tissue. Do the same
for theirs. If the list doesn't give it away, explain why it
challenging.

- They said it was hard to determine which tissue these genes belong to
  because they are all mitochondrial genes. They guessed that maybe it
  was the heart since the heart needs a lot of energy.

**(c)** Why is the `> 10 TPM` filter in part (a)? Take it out and look
at what rises to the top. What kind of genes are the most \`\`specific"
when you do not require them to be meaningfully expressed?

```{r}
dt %>%
  mutate(specificity = log2((`Brain - Amygdala`+1)/(mean(mean_all_tissues)+1))) %>%
  select(Description,specificity, `Brain - Amygdala`) %>%
  arrange(-specificity) %>%
  select(Description) %>%
  head(20)

#there was no change in the 20 genes with the highest specificity, so it appears the amygdala specificity score is not affected by genes with a TPM lower than 10.

```

**(d)** Housekeeping genes, which are highly expressed in every tissue,
should score near 0 on this specificity metric. Find a few genes with
specificity closest to 0 and high expression. Do they look like what you
would expect?

```{r}
dt %>%
  mutate(specificity = log2((`Brain - Amygdala`+1)/(mean(mean_all_tissues)+1))) %>%
  select(Description,specificity, `Brain - Amygdala`) %>%
  arrange(abs(specificity))

#I didn't see anything that I recognized to be a housekeeping gene
```

------------------------------------------------------------------------

### Q6. Comparing gene expression between tissues

Another way to compare tissues is to ask how similar are their
expression patterns overall, across all genes?

R has a built-in function called `cor.test()` that measures the
correlation between two vectors of numbers. Remember that you can
extract the contents of a given column of a tibble as a vector using the
syntax `tibble_name$column_name`. If a column name contains spaces,
enclose the column name in backticks such as
`` tibble_name$`column name` ``.

Use `cor.test()` with the argument `method = "spearman"` to compute:

- the Spearman correlation between `Liver` and `Heart - Left Ventricle`
- the Spearman correlation between `Liver` and `Brain - Cortex`

Which pair of tissues is more strongly correlated? Does this result make
sense biologically?

You will see a warning about not being able to compute an exact p-value
because of ties. You can ignore it, as it does not affect the
qualitative results in this case. It is worth understanding where it
comes from, though: what is it about these data that creates so many
tied values?

```{r q6}
liver <- dt$Liver
heart <- dt$`Heart - Left Ventricle`
brain <- dt$`Brain - Cortex`
# Liver vs Heart answer
cor.test(liver, heart, method = "spearman")
# Liver vs Brain answer
cor.test(liver, brain, method = "spearman")
# it appears that liver and heart are more strongly correlated than the brain. This might make sense because the brain is a different organ compared to liver and heart.
```

------------------------------------------------------------------------

### Optional: further exploration

What other patterns are present in the data? Use tidyverse functions to
explore and summarize the data in different ways and see what you can
learn.

```{r optional}
# answer here
```
